---
title: "Databases and modeling with R"
author: "Alex Farach"
date: "3/21/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(odbc)
library(DBI)
library(RMySQL)
library(dbplyr)
```

# Create connection

```{r}
drv <- dbDriver("MySQL")

con <- dbConnect(
  drv,
  user = 'admin',
  password = 'mydbinstance',
  dbname = 'mydb',
  host = 'mydbinstance.c6c1szwymy7m.us-east-1.rds.amazonaws.com',
  port = 3306
)
```

## Create table for use in example

```{r}
DBI::dbListTables(con)
```
No tables. Let's create one and drop it in.

The following table comes from the kaggle dataset: https://www.kaggle.com/ntnu-testimon/paysim1

```{r}
fraud <- data.table::fread("../FRTIB/fraud/example_data/fraud_example.csv")
```

```{r}
dbWriteTable(con, "fraud_example", fraud)
```

## Let's get started

What tables do we have in our data lake?
```{r}
dbListTables(con)
```
Looks like we only have 1 table "fraud_example". Let's look at the column names
in this table.

```{r}
dbColumnInfo(con, "fraud_example")
```

Cool. We can also use these 2 commands to create a function that creates a table
showing all of the tables and the columns within them.

```{r}
list_of_tables <- dbListTables(con)
db_complete <- NULL

for (i in list_of_tables) {
  loop_db <- dbColumnInfo(con, "fraud_example") %>%
    mutate(
      table_name = "fraud_example"
    )
  
  db_complete = bind_rows(db_complete, loop_db)
}

db_complete
```

# Multiple ways to get data

You can write SQL and send/get results using `r DBI::dbGetQuery()`
```{r}
fraud_data <- dbGetQuery(con, "
                         SELECT * 
                         FROM fraud_example 
                         ORDER BY RAND() 
                         LIMIT 25000;")
```

Or use actual SQL and the RMarkdown sql engine
```{sql connection = con, output.vars = "fraud_data"}
SELECT * 
FROM fraud_example 
ORDER BY RAND() 
LIMIT 25000;
```

Or avoid SQL all together and use `r dbplyr()` which converts `dplyr()` syntax 
into `SQL`.
```{r}
# create tbl() of the table you need to query
fraud <- tbl(con, "fraud_example")

# write dplyr like you normally would. dbplyr does the rest
fraud_data_summary <- fraud %>%
  select(-row_names) %>%
  group_by(isFraud, isFlaggedFraud, step) %>%
  summarise(
    n = n()
  ) %>%
  ungroup() %>%
  # select show_query() to see the translation dbplyr is sending along
  #show_query()
  # select collect() to collect the results of the query and bring into R
  collect()
```

We now have a dataset we can take a look at.

```{r}
fraud_data_summary %>%
  group_by(isFraud, isFlaggedFraud) %>%
  summarise(
    n = n()
  ) %>%
  mutate(prop = scales::percent(prop.table(n), accuracy = 0.001))

fraud_data_summary %>%
  group_by(step) %>%
  summarise(
    n = sum(n)
  ) %>%
  mutate(prop = scales::percent(prop.table(n), accuracy = 0.001))

fraud_data %>%
  select(-row_names) %>%
  group_by(isFraud) %>%
  summarise(
    n = n()
  ) %>%
  ungroup()

fraud_data %>%
  select(-row_names) %>%
  count(isFraud) %>%
  mutate(per_n = scales::percent(n/sum(n), accuracy = 0.01))
```


