---
title: "dss2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(here)
```


# Data science social 2

## import data
```{r}
oes <- read.csv("./oes_2019.csv") %>%
  janitor::clean_names(.)

glimpse(oes)
```

I want to find the data science related fields

```{r}
oes %>%
  mutate(occ_title = tolower(occ_title)) %>%
  filter(str_detect(occ_title, "data")) 
```

```{r}
oes %>%
  mutate(occ_title = tolower(occ_title)) %>%
  filter(str_detect(occ_title, "data")) %>%
  distinct(occ_title)
```

This looks promising. I'll start here. It might be worthwhile to come back at
some point and look through all the different tech occupations.

I'm going to filter by occupation code.
```{r}
oes %>%
  filter(occ_code == "15-2098") %>%
  select(occ_code, o_group, i_group, everything())
```
I see different levels of things like o_group and i_group. Also area and naics
looks like it needs to be cleaned up. Area is the FIPS code and should be a factor
2 characters long. Naics is all over the place. Sometimes its 6 digits long, sometimes
2, sometimes it's hyphenated. I need to figure out how to clean that up if I want
to use it.

I didn't see this before but it looks like the expected numeric variables like 
pct_total or a_mean are all character! What the heck? I'm pretty sure it's because
of those asteriks and other special characters in there. We need to get rid of 
those in order to set these variables as numeric an be able to use.

Lots of cleanup...first things first though, let me figure out what NAICS should
look like

```{r}
library(searcher)

search_google("bls naics code structure", rlang = FALSE)
```
Actually I have another idea. There has to be a relationship between o_group, i_group
and the structure of the naics code.
```{r}
oes %>%
  filter(occ_code == "15-2098") %>%
  mutate(naics_length = nchar(naics)) %>%
  distinct(o_group, i_group, naics_length) %>%
  arrange(desc(i_group), o_group)
```
Ok, what this tells me is that the occ code for data scientists (15-2098) can 
exist in different o_groups and i_groups. I need to learn more about those to
figure out what exactly I should be looking at.

## industry code structure

```{r}
search_google("oes 'i_group'", rlang = FALSE) # that didn't help
search_google("oes data structure", rlang = FALSE)
```
I see now. i_group tells us where in the heirarchy the occupation code is in. The 
farther down the heirarchy, with the 6-digit level being the farthest down, the more
specific the data is.

```{r}
search_google("oes detailed major total", rlang = FALSE)
```
Ok, now we're talking. The second link "Technical Notes for May 2019 OES Estimates"
has a section for "The occupational coding system" and "The industry coding system", 
presumably o_group and i_group. Best of all it has links to the specific pages
where information about these groups can be found. 

After some clicking I got here: https://www.bls.gov/soc/2018/soc_structure_2018.pdf.
Perfect.

The hyphen indicates a combo. So naics 31-33 means that the values presented are
for 310000 and 330000 combined. That's kind of a headache to deal with :-/. How often
does that happen?

```{r}
oes %>%
  filter(str_detect(naics, "-"),
         occ_code == "15-2098") %>%
  distinct(naics)
```

2 times. I think we can handle this.

## occupation code structure

To get a sense of what the occupation group structure looks like we can do something
like:

```{r}
oes %>%
  filter(occ_code == "15-2098") %>%
  mutate(
    occ_code = str_remove_all(occ_code, "0"),
    occ_code = str_remove_all(occ_code, "-"),
    o_group_length = nchar(occ_code)
    ) %>%
  distinct(o_group, o_group_length)
```

Ok, cool. So the data science occupation code only exists at the 6 occupation code
level. This means we don't have to worry about parsing through the other 2 categories:
total and major.

So now we know that the data scientist occupation of 15-2098 exists in different
industries at different levels (sector, 3, 4, 5, 6 digit level) but only at the 
6 digit occupation level. We can remove the o_group then so long as we always
filter for the 15-2098.

We still have to clean up some more but we're getting there.

```{r}
oes_clean_ds <- oes %>%
  filter(occ_code == "15-2098") %>%
  mutate_all(., ~ str_replace(., "\\**", ""),
             ., ~ str_replace(., "~", "")) %>%
  mutate(
    fips = as.factor(str_pad(area, 2, "left", "0")),
    naics = case_when(
      naics == "31-33" ~ "310000",
      naics == "44-45" ~ "440000",
      TRUE ~ naics
    ),
    naics = as.factor(str_pad(naics, 6, "right", "0")),
    area_title = as.factor(tolower(area_title)),
    naics_title = tolower(naics_title),
    occ_title = tolower(occ_title),
    i_group_digit = if_else(i_group == "sector", "2", str_sub(i_group, 1, 1)),
    across(c("tot_emp":"hourly"), as.numeric)
  ) %>%
  group_by(fips, area_title) %>%
  mutate(max_detail = max(i_group_digit)) %>%
  filter(i_group_digit == max_detail) %>%
  select(fips, area_title, i_group, i_group_digit, max_detail, naics, naics_title, starts_with("a_")) %>%
  arrange(fips, naics)

oes_clean_ds
```
Time to plot!

```{r}
oes_clean_ds %>%
  ungroup() %>%
  mutate(naics_2 = as.factor(paste(area_title, naics_title, sep = ": ")),
         naics_2 = str_to_title(naics_2),
         naics_2 = fct_reorder(naics_2, a_mean)) %>%
  ggplot(aes(x = naics_2, y = a_mean)) +
  geom_segment(aes(x=naics_2, xend=naics_2, y=0, yend=a_mean), linetype = "dashed") +
  geom_point( size = 4, color="orange", alpha = 0.6) +
  coord_flip() +
  ggthemes::theme_fivethirtyeight() +
  labs(
    title = "Data Scientists and Mathematical Science Occupations \nAverage Salary by State and Industry",
    x = NULL,
    y = NULL,
    caption = "Source: BLS Occupational Employment Statistics \nhttps://www.bls.gov/oes/home.htm"
  ) +
  theme(
    plot.title = element_text(hjust = 1),
    panel.grid.major.y = element_blank()
  ) +
  scale_y_continuous(labels = scales::dollar)

```
